Emittance preservation of an electron beam in a loaded quasi-linear plasma wakefield

### 2017-10-05 Tex/Paper.tex
Changes from Erik and Patric

\documentclass[aps,prstab,reprint,amsmath,amssymb,groupedaddress]{revtex4-1}
\bibliographystyle{apsrev4-1}

@@

@@

\begin{abstract}
We investigate beam loading and emittance preservation for a high-charge electron beam being accelerated in quasi-linear
plasma wakefield driven by a short proton beam. The structure of the wakefield is similar to that of a long, modulated
proton beam, such as the one being studied by AWAKE.  We show that by exploiting two well known effects, full blow out
of plasma electrons by the accelerated beam, and beam loading of the wake field, the electron beam can be accelerated
without significant emittance growth.
\end{abstract}

@@

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades \cite{chen:1985}. With a relativistic charged particle drive
beam travelling through a plasma, a strong wakefield is excited, that can be loaded by a trailing witness beam. When the
witness beam optimally loads the wakefield, an increase in absolute energy spread can be kept to a minimum. The concept
has been demonstrated experimentally in the past using electron drive beams accelerating electron witness beams
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008}. AWAKE at CERN is a proof of concept beam driven plasma wakefield
accelerator experiment using a proton drive beam delivered by the SPS accelerator \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to accelerate a beam while keeping energy spread and
emittance growth to a minimum. In the well described linear case where the beam density $n_{b}$ is much smaller than the
plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the witness beam. the beam
will also see a transversely and longitudinally varying accelerating field causing a spread in energy after the beam has
been accelerated \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by the
transverse oscillations of the plasma electrons, forming a sheet around an evacuated area filled with only ions. The
ions, assumed stationary, form an uniform density ion channel creating a focusing force that varies linearly with
radius. This produces a focusing force that preserves emittance \cite{lu:2006-1, lu:2006}.

In this paper we present simulation results showing how we can use the non-linear wakefields driven by the head of a
high charge density witness beam to reduce energy spread and emittance growth in the rest of the beam, while being
accelerated by a proton drive beam producing quasi-linear wakefields.

% ******************************************************************************************************************** %
\subsection[\label{S:I:SMI}]{Self-modulation as a Driver}
%~ Patric's Comments
%~ - Self-modulation as a driver
%~ - SM does not reach blow out
%~ - SM does not preserve emittance and produces large energy spread and parameters may evolve along the accelerator,
%~   not desirable
% ******************************************************************************************************************** %

A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where the
plasma wavelength $\lambda_{pe} = 2\pi c/\omega_{pe}$ and the plasma frequency $\omega_{pe} = (n_{0} e^{2} / m_{e}
\epsilon_{0})^{1/2}$, produces a field $E_{z}$ that increases for each drive bunch \cite{chen:1985}. A tailing witness
beam loading the peak accelerating phase of this field will gain energy from the wakefield from all of the drive bunches
when they are properly phased with respect to the wakefields. Acceleration of an electron witness beam driven by two
electron drive beams has been demonstrated at Brookhaven National Laboratory \cite{muggli:2011}.

The energy carried by electron drive bunches used in previous experiments were typically small -- on the order of
$100\unit{J}$ -- and the propagation length typically $<1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. Higher energy
electron beam can be improve the propagation length and thus the energy of the accelerated beam. For instance, the
energy of a high-charge electron beam accelerated to $1\unit{TeV}$, similar to the beam that could be produced by the
international linear collider with $1\nexp{10}$ electrons, is $1.6\unit{kJ}$. By using electron beams or lasers pulses
as drivers, a large number of plasma stages is required. However, staging plasma accelerators without reducing the
effective gradient and spoiling the beam quality is challenging \cite{steinke:2016,lindstrom:2016}.

Proton beams available at CERN carry significantly more energy, $19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016},
allowing for much longer plasma wakefield accelerator stages. The SPS beam is, unfortunately, orders of magnitude longer
than the plasma wavelengths needed for such applications. This can be resolved by letting the proton beam undergo
self-modulation before injecting the witness beam into accelerating structure. The self-modulation of the beam is
produced by the transverse fields generated by the beam acting upon itself, causing regions of the beam to rapidly
defocus \cite{kumar:2010}. The modulation frequency is close to that of the plasma, and produces a train of short proton
bunches along the beam axis with a surrounding halo of defocused particles. This train of bunches resonantly produces
wakefields to larger amplitudes.

@@

The AWAKE experiment, currently in its first stages of operation at CERN, uses a $400\unit{GeV}$ beam delivered by the
SPS as its driver, and a single $10\unit{m}$ plasma stage with a nominal plasma density of $7\nexp{14}\unit{cm}^{-3}$
\cite{gschwendtner:2016}. This plasma density corresponds to $\lambda_{pe} = 1.26\unit{mm}$ -- matched to the transverse
size of the SPS proton beam such that $k_{pe}\sigma_{x,y} = 1$, where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave
number.

The aim of the first phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam. The aim of
the second phase in 2018 is to sample the wakefield with a long electron beam $\simeq\lambda_{pe}$. The study presented
here is for Run 2 \cite{adli:2016}, which aims to demonstrate acceleration of a short electron beam, $\ll\lambda_{pe}$,
to high energy with a minimal increase in emittance and absolute energy spread.

\begin{figure}[hbt]
    \includegraphics[width=0.99\linewidth,trim={1mm 2mm 1mm 2mm},clip]{figures/figAWAKE}
    \caption{\label{Fig:AWAKER2} A simplified illustration of the experimental setup for AWAKE Run 2. The SPS proton
        beam undergoes self-modulation in the first plasma section. The electron witness beam is injected into the
        accelerating structure, and undergoes acceleration in the second plasma stage
        \cite{berglyd_olsen:2015, adli:2016}.}
\end{figure}

The plans for AWAKE Run 2 proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}. The first
section of $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without the electron
beam present. The electron witness beam is then injected into the modulated proton beam before stage two, where it
undergoes acceleration.

Using the SPS proton beam as driver, the self-modulated proton beam does not produce a full non-linear wakefield, and
not all plasma electrons are therefore evacuated from the plasma bubble. The result is that the focusing force does not
increase linearly with radius. The small size of the accelerating structure produced by the self-modulated beam also
puts constraints on the size of the electron witness beam, and thus its charge and current, if at the same time we wish
to prevent large energy spread. By matching the transverse size of the beam at a given emittance to the plasma density,
we prevent large amplitude oscillations which may cause additional energy spread as well as emittance growth.

%~ Eric's version:
%~ The key idea in this paper is that while the proton beam wake is not fully blown-out, and thus leading to non-linear
%~ focusing forces, the electron beam, if intense enough, may load the field and at the same time create its own ion
%~ bubble. As will be shown, exploiting these two well known effects, the electron beam can be accelerated for long
%~ distances without significant emittance growth.

%~ Patric's version:
The key idea is to have enough charge in the witness beam to at the same time load the wakefield to produce low
relative energy spread $\Delta E/E$, and blow out the electrons left in the accelerating structure to reach conditions
that preserves emittance.

@@

A main focus of this study is on the loading of the wakefields driven by the proton beam. In order to eliminate other
factors that may affect this, like decay and defocusing of the proton beam, we tried several approaches to create a
stable drive beam structure based in previous self-modulation studies.

In these studies we used a short, pre-modulated proton beam with similar structure to the beam produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations and primarily looked at beam loading, energy gain and energy spread
\cite{berglyd_olsen:2015, berglyd_olsen:2016}.

In order to study the witness beam emittance evolution we turned to the recently released open source version of
QuickPIC. QuickPIC is a fully relativistic 3D quasi-static PIC code \cite{huang:2006, an:2013}. It does not suffer from
the numerical Cherenkov effect that full PIC codes do \cite{godfrey:1974,lehe:2013}, making it a well suited tool to
study emittance preservation.

@@

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/plasmaDenTWake}
    \caption{\label{Fig:PlasmaDenTWake} The plasma density in grey with the drive beam (blue) and the witness beam (red)
        superimposed. The line plot indicates the transverse wakefield gradient $dW_{x}/dx$ where
        $W_{x} = E_{x} - v_{b} B_{y}$, evaluated along the beam axis.}
\end{figure}

@@

The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. When
the SPS beam, containing $3\nexp{11}$ protons \cite{gschwendtner:2016}, enters the second plasma stage, the peak
electric field is expected to be $500-600\funit{MV}{m}$. The plasma electrons are only depleted to around $65\%$ of
nominal plasma density at the point where we inject the electron beam \cite{awake_collaboration:2016}. In our simulation
these conditions are replicated using a single proton beam of $1.46\nexp{10}$ protons, or $2.34\unit{nC}$ and
$7\unit{kA}$.

The peak density of the simulation drive beam is $0.83\cdot n_{0}$, producing a quasi-linear wakefield. The single bunch
setup uses the baseline proton beam energy of $400\unit{GeV}$ and transverse size $\sigma_{x} = 200\unit{\mu m}$. We
also set the length of the drive beam to $\sigma_{z} = 40\unit{\mu m}$. We are not considering the evolution of the
proton beam itself in this study, so in order to create a stable environment to study the evolution of the witness beam
we prevent the proton beam from evolving radially by increasing the particle mass by 6 orders of magnitude.

@@

Beam loading of a short witness beam is sensitive to its position relative to the electric field \cite{tzoufras:2009}.
A low gamma witness beam can easily drift out of phase with the drive beam wakefield. The drift of the beams is
proportional to $\gamma^{-2}$, so to eliminate initial de-phasing of the witness beam, initial beam energy was set such
that $\gamma_{eb} = \gamma_{pb} = 426.3$, giving an energy of $217\unit{MeV}$. A lower initial value is likely to be
sufficient \cite{berglyd_olsen:2015}.

In these simulations we have used a matched beam for an initial normalised emittance of $\epsilon_{N} = 2\unit{\mu m}$.
The beam matching relation is given by
\begin{equation}
    \beta = \frac{\sigma_{x,y}^{2}\gamma}{\epsilon_{N}}
          = \frac{c}{\omega_{pe}}\sqrt{2\gamma}, \label{EQ:MatchedB}
    % = c\sqrt{\frac{m_{e}\varepsilon_{0}}{n_{pe}e^{2}}2\gamma}
\end{equation}
where $\beta$ is the Twiss parameter. It then follows that the transverse profile of the beam must satisfy
\begin{equation}
    \sigma_{x,y}^{2} = c\epsilon_{N}\sqrt{2\frac{m_{e}\varepsilon_{0}}{n_{pe}e^{2}\gamma}} \label{EQ:MatchedR}
\end{equation}
in order to be matched to a given plasma density, initial emittance and energy. This relation requires a very narrow
beam with a $\sigma_{x,y}$ of $5.25\unit{\mu m}$ compared to the drive beam $\sigma_{x,y} = 200\unit{\mu m}$. The charge
density of this compact witness beam reaches that of the plasma density at only a few $\unit{pC}$, but reaches optimal
beam loading at around $100-200\unit{pC}$. The implication here being that the witness beam produces its own non-linear
wakefield already at the head of the beam. The majority of the electrons within it will therefore see a linear focusing
force preventing the bulk of the beam from undergoing emittance growth.

The relatively small size of the witness beam compared to the proton beam put some restrictions on the transverse
resolution of the witness beam as we need a simulation box wide enough to contain the larger beam while resolving the
narrower one (see figure \ref{Fig:PlasmaDenTWake}). We used a transverse grid cell size of $1.17\unit{\mu m}$, and of
$2.34\unit{\mu m}$ for the longitudinal grid cells for the simulations presented in section \ref{S:BL}. The witness beam
was simulated with $16.8\nexp{6}$ non-weighted particles. The witness beam has a total charge of $100\unit{pC}$, a
$\sigma_{z}=60\unit{\mu m}$, and a $\sigma_{x}=5.25\unit{\mu m}$ matching an initial normalised emittance
$\epsilon_{0} = 2\unit{\mu m}$. We refer to this parameter set as our base case.

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamLoading}
    \caption{\label{Fig:BeamLoading} The top plot shows the unloaded longitudinal electric field with no witness beam
        (dashed blue line) and the loaded field (whole blue line) along the beam axis. The beam density along the axis
        for both beams are shown in red. The bottom plot compares the plasma densities along the beam axis for a drive
        beam with no witness beam (dashed green line), witness beam with no drive beam (dash-dotted green line), and
        both beams present (whole green line). $\xi = z - tc$ is the position in the simulation box, moving towards the
        left.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamPhaseSpace}
    \caption{\label{Fig:BeamPS} The phase space charge distribution of a $100\unit{pC}$, $60\unit{\mu m}$ long witness
        beam after $4\unit{m}$ of plasma. The mean momentum is $1.67\unit{GeV/c}$ with an RMS energy spread of
        $87\unit{MeV/c}$ ($5.2\%$) for the full beam.}
\end{figure}

@@

\begin{figure*}[hbt]
    \includegraphics[width=\linewidth,trim={0mm 0mm 0mm 0mm},clip]{figures/beamFilamentationAll}
    \caption{\label{Fig:BeamFilament} Plots \textbf{a} to \textbf{f} show the transverse phase space of the offset
        electron beam at different plasma positions. Plot \textbf{g} shows the macro particle mean position, and plot
        \textbf{h} their RMS spread. Plots \textbf{a}, \textbf{b} and \textbf{c}, as well as the blue lines in plots
        \textbf{g} and \textbf{h} represent particles with position $1.40\unit{\mu m} < \xi < 1.42\unit{\mu m}$. Plots
        \textbf{d}, \textbf{e} and \textbf{f}, and the red lines in plots \textbf{g} and \textbf{h} represent particles
        with position $1.55\unit{\mu m} < \xi < 1.57\unit{\mu m}$.}
\end{figure*}

@@

The simulations for this study have been run using the open source version of QuickPIC released in early 2017 and owned
by UCLA.

>>>>

\documentclass[aps,prstab,reprint,amsmath,amssymb,groupedaddress,onecolumn]{revtex4-1}
\bibliographystyle{apsrev4-1}

@@

%Patric general comments
%energy labels not consistent between figs 7, 9 and 8
%labels for the fractional charge is too complicated on Figs 7,8,9. it needs to be explained in the caption and/or in the text. "On the left axis we plot the fraction of that charge whose emittance remains .... "
%Add scale for grey color on Fig. 2
%Fig. 3 has a "v=c" and an arrow that are only on that figure ...
%It is not clear to me what the conclusions/outcome of the last section are ....

@@

\begin{abstract}
We investigate beam loading and emittance preservation for a high-charge electron beam being accelerated in quasi-linear
plasma wakefield driven by a short proton beam. The structure of the studied wakefields are similar to those of a long, modulated
proton beam, such as the AWAKE proton driver.  We show that by %
properly choosing the electron beam parameters and exploiting two well known effects, beam loading of the wake field and full blow out
of plasma electrons by the accelerated beam, the electron beam can %PM be accelerated - EA ok
gain large amounts of energy with a narrow final energy spread (\%-level) and without significant emittance growth.
\end{abstract}

@@

Beam driven plasma wakefield accelerators \cite{chen:1985} have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades. With a relativistic charged particle drive
beam travelling through a plasma, a strong wakefield is excited, that can be loaded by a trailing witness beam. When the
witness beam optimally loads the wakefield, an increase in absolute energy spread can be kept small. %to a minimum. %PM - EA ok
The concept
has been demonstrated experimentally in the past using electron drive beams accelerating electron witness beams
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008} [EA: M. Litos et al., Nature 515, 92?95 (2014)].

A major challenge with plasma wakefield accelerators is, however, to accelerate a beam while keeping both energy spread and
emittance growth small. %growth to a minimum. PM - EA ok
In the well described linear regime, valid when the beam density $n_{b}$ is much smaller than the
plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the witness beam. The beam
%will %PM - EA ok
also sees a transversely and longitudinally varying accelerating field causing a spread in energy after the beam has
been accelerated \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by the
transverse oscillations of the plasma electrons, %forming PM - EA ok
gathering in a sheet around an evacuated area filled with only ions. The
ions, assumed stationary, form an uniform density ion channel creating a focusing force that varies linearly with
radius. This %produces a  PM - EA ok
focusing force %that PM - EA ok
preserves emittance \cite{rosenzweig:1991}.
% EA: I put Rosenzweig original blow out paper as citation here instead of, \cite{lu:2006-1, lu:2006}.

In this paper we present simulation results showing how emittance preservation of a high charge density witness beam can be ensured, when accelerated by a proton drive beam producing quasi-linear wakefields \cite{rosenzweig:2010}.  By quasi-linear wakefields we here mean wakefields with only partial blow out of the plasma electrons in the bubble.
%~ Patric's version is implemented:
The key idea is to have enough charge in the witness beam to at the same time load the wakefield to produce low
relative energy spread%$\Delta E/E$, 
, and completely blow out the electrons left in the bubble after the beam to reach conditions that preserves emittance.
The results have importance for the preparation of AWAKE Run 2 and possibly other applications in the quasi-linear regime.

% ******************************************************************************************************************** %
%\subsection[\label{S:I:SMI}]{Self-modulation as a Driver}
%~ Patric's Comments
%~ - Self-modulation as a driver
%~ - SM does not reach blow out
%~ - SM does not preserve emittance and produces large energy spread and parameters may evolve along the accelerator,
%~   not desirable
% ******************************************************************************************************************** %

% 	EA comment: is this part relevant?  Feels out of place.  We repeat this physics below when discussing proton beam SMI.  
%A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where the
%plasma wavelength $\lambda_{pe} = 2\pi c/\omega_{pe}$ and the plasma frequency $\omega_{pe} = (n_{0} e^{2} / m_{e}
%\epsilon_{0})^{1/2}$, produces a field $E_{z}$ that increases for each drive bunch \cite{chen:1985}. A tailing witness
%beam loading the peak accelerating phase of this field will gain energy from the wakefield from all of the drive bunches
%when they are properly phased with respect to the wakefields. Acceleration of an electron witness beam driven by two
%electron drive beams has been demonstrated at Brookhaven National Laboratory \cite{muggli:2011}.

@@

The energy carried by electron drive bunches used in previous plasma wakefield experiments have been no more than
$100\unit{J}$ and the propagation length typically no more than $1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}.  For high-energy physics application a higher total beam energy is often desired, for instance, the energy of a high-charge electron bunch accelerated to $1\unit{TeV}$ with $1\nexp{10}$ electrons, similar to the beam that could be produced by the
international linear collider, is $1.6\unit{kJ}$.  Using electron beams 
as drivers a large number of plasma stages is required to reach %PM - EA ok
an energy of a $\unit{kJ}$ for the accelerated bunch. However, staging plasma accelerators without reducing the
effective gradient and spoiling the beam quality is challenging \cite{steinke:2016,lindstrom:2016}.

Proton beams available at CERN carry significantly more energy than available electron beams, $19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016},
allowing for much longer plasma wakefield accelerator stages. The SPS beam is orders of magnitude longer
than the plasma wavelengths needed for such applications, and the beam %will PM - EA ok
does not drive a strong wake \cite{gschwendtner:2016}.  By letting the proton beam undergo
self-modulation before injecting the witness beam into accelerating structure, %most PM - EA ok
stronger wakefields are %formed PM - EA ok
excited. The self-modulation of the beam is
produced by the transverse fields generated by the beam acting upon itself, causing regions of the beam to rapidly
defocus \cite{kumar:2010}. The modulation frequency is close to that of the plasma, and produces a train of short proton
bunches along the beam axis with a surrounding halo of defocused particles. This train of bunches resonantly %produces - EA ok
drives
wakefields to large %r PM - EA ok 
amplitudes.

AWAKE at CERN is a proof of concept proton beam driven plasma wakefield
accelerator experiment \cite{awake_collaboration:2014}, currently in its first phase of operation. The experiment uses a $400\unit{GeV}$ proton beam delivered by the
SPS as its driver, and a single $10\unit{m}$ plasma stage with a nominal plasma density of $7\nexp{14}\unit{cm}^{-3}$
\cite{gschwendtner:2016}. This plasma density corresponds to $\lambda_{pe} = 1.26\unit{mm}$ -- matched to the transverse
size of the SPS proton beam such that $k_{pe}\sigma_{x,y} = 1$, where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave
number, %PM. - EA ok
$\lambda_{pe}=2\pi c/\omega_{pe}$ and $\omega_{pe}=\left(n_0e^2/m_e\epsilon_0\right)^{1/2}$ is the plasma electron angular frequency. %

The aim of the first phase of the experiment is to demonstrate self-modulation of the proton beam. The aim of
the second phase, in 2018, is to sample the wakefield with a long electron beam $\simeq\lambda_{pe}$. The study presented
here has relevance for Run 2 \cite{adli:2016}, starting in 2021 after the LHC long shut down 2, and aims to demonstrate acceleration of a short electron beam, $\ll\lambda_{pe}$,
to high energy and with a minimal increase in emittance and absolute energy spread.

\begin{figure}[hbt]
    \includegraphics[width=0.99\linewidth,trim={1mm 2mm 1mm 2mm},clip]{figures/figAWAKE}
    \caption{\label{Fig:AWAKER2} A simplified illustration of the experimental setup for AWAKE Run 2. The SPS proton
        beam undergoes self-modulation in the first plasma section. The electron witness beam is injected into the
        accelerating structure, and undergoes acceleration in the second plasma section
        \cite{berglyd_olsen:2015, adli:2016}.}
\end{figure}

The plans for AWAKE Run 2 proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}. The first
section of about $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without the electron
beam present. The electron witness beam is then injected into the modulated proton beam before section two, where it
undergoes acceleration. The self-modulated proton beam does not produce a fully non-linear wakefield, and
 therefore not all plasma electrons are evacuated from the plasma bubble. The result is that the focusing force does not
increase linearly with radius %PM
and the accelerated beam emittance is not preserved.

%   EA: Moved to "witness beam"
%The small size of the accelerating structure produced by the self-modulated beam also
%puts constraints on the size of the electron witness beam, and thus its charge and current, if at the same time we wish
%to prevent large energy spread. By matching the transverse size of the beam at a given emittance to the plasma density,
%we prevent large amplitude oscillations which may cause additional energy spread as well as emittance growth.

%    EA: I kept Patric's version, but moved it to the end of the introduction
%~ Eric's version:
%~ The key idea in this paper is that while the proton beam wake is not fully blown-out, and thus leading to non-linear
%~ focusing forces, the electron beam, if intense enough, may load the field and at the same time create its own ion
%~ bubble. As will be shown, exploiting these two well known effects, the electron beam can be accelerated for long
%~ distances without significant emittance growth.

@@

The focus in this study is the beam loading of the wakefields driven by the proton beam.   Studies of self-modulated proton bunches show that the proton bunch evolve as it propagates through a uniform plasma [EA: K. V. Lotov, Phys. Plas. 18, 024501 (2011], but that small variations in the plasma profile the modulation, and thus the wake fields, may be stabilized over long distances [EA: K. V. Lotov, Phys. Plas. 18, 024501 (2011), K. V. Lotov, Phys. Plas. 22, 103110 (2015), A. Caldwell and K. Lotov, Phys. Plas. 5, 103101 (2011)].  To study the witness beam evolution in a stable wake, independent of the dynamics of the self-modulation, %PMwe define here - EA ok
here we use a single, non-evolving proton bunch as driver.  The proton bunch parameters are chosen so that key features of the wake - the plasma electron density in the wake and the longitudinal electric field - are the same as %PM - EA ok
in the wake of a self-modulated proton beam with AWAKE baseline parameters \cite{gschwendtner:2016}.  Both the proton beam and the witness have Gaussian %PM - EA ok
longitudinal charge distribution and bi-Gaussian transverse charge distributions.

We have previously studied the beam loading in a proton beam wake using the full PIC code Osiris \cite{fonseca:2002}
%using PM - EA ok
and with 2D cylindrical-symmetric simulations.  The studies \cite{berglyd_olsen:2015, berglyd_olsen:2016} primarily looked at beam loading, energy gain and energy spread, as well as different approaches to create a stable drive beam structure based on previous self-modulation studies.  In order to study the witness beam emittance evolution we adopt the recently released open source version of QuickPIC \cite{huang:2006, an:2013}. QuickPIC is a fully relativistic 3D quasi-static PIC code. It does not suffer from
the numerical Cherenkov effect that full PIC codes do \cite{godfrey:1974,lehe:2013}, making it a well suited tool to
study emittance preservation.  All simulation results in this paper are %PMcalculated - EA ok
obtained using QuickPIC open source.

@@

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/plasmaDenTWake}
    \caption{\label{Fig:PlasmaDenTWake} %PMThe p
    Plasma density in grey with the drive beam (blue) and the witness beam (red)
        superimposed. The line plot indicates the transverse wakefield gradient $dW_{x}/dx$ where
        $W_{x} = E_{x} - v_{b} B_{y}$, evaluated along the beam axis.
        %PM
        Beams move to the left.}
\end{figure}

@@

%PM The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. - EA ok
The modulation process used in AWAKE does not reach the fully non-linear regime and thus does not produce a bubble void of plasma electrons. When
the SPS beam, containing $3\nexp{11}$ protons \cite{gschwendtner:2016}, enters the second plasma stage (figure \ref{Fig:AWAKER2}), the peak
electric field is expected to be about $500\funit{MV}{m}$. The plasma electron density are only depleted to around $65\%$ of
nominal plasma density at the point where we inject the electron beam \cite{awake_collaboration:2016}.  The plasma electron density depletion and the peak field is replicated using a single %PM proton 
bunch with $1.46\nexp{10}$ protons (or $2.34\unit{nC}$), a %PM bunch
length $\sigma_{z} = 40\unit{\mu m}$ ($7\unit{kA}$), and a transverse size $\sigma_{x,y,proton} = 200\unit{\mu m}$.   The peak density is $0.83\cdot n_{0}$ results in a quasi-linear wake. %PM The beam energy is $400\unit{GeV}$.  
To avoid transverse evolution of the proton driver, emulating the stable propagation of the self-modulated proton beam [EA: K. V. Lotov, Phys. Plas. 18, 024501 (2011), K. V. Lotov, Phys. Plas. 22, 103110 (2015), A. Caldwell and K. Lotov, Phys. Plas. 5, 103101 (2011)] we freeze the transverse evolution of the equivalent proton bunch by increasing the particles mass by six orders of magnitude.

@@

%The small size of the accelerating structure produced by the self-modulated beam also
%puts constraints on the size of the electron witness beam, and thus its charge and current, if at the same time we wish
%to prevent large energy spread.  
%PM not sure what the point of the above sentence is ... - EA ok
In order to prevent large amplitude oscillations %PM
of the witness beam particles, which may cause additional energy spread as well as emittance growth, we %PMstudy
consider a witness beam matched to the plasma density.  
%The beam matching relation is given by
%\begin{equation}
%    \beta = \frac{\sigma_{x,y}^{2}\gamma}{\epsilon_{N}}
%          = \frac{c}{\omega_{pe}}\sqrt{2\gamma}, \label{EQ:MatchedB}
%    % = c\sqrt{\frac{m_{e}\varepsilon_{0}}{n_{pe}e^{2}}2\gamma}
%\end{equation}
%where $\beta$ is the Twiss parameter. It then follows that the transverse profile of the beam must satisfy
%PM Not sure what the point of the above lines is ....
The matched beam transverse size is:
\begin{equation}
%    \sigma_{x,y}^{2} = c\epsilon_{N}\sqrt{2\frac{m_{e}\varepsilon_{0}}{n_{pe}e^{2}\gamma}} \label{EQ:MatchedR}
%    \sigma_{x,y} =\left(c^2\epsilon_{N}^2m_{e}\varepsilon_{0}n_{pe}e^{2}\gamma \right)^{1/4} \label{EQ:MatchedR} - % PM - EA: not correct
	\sigma_{x,y}=\left(\frac{2c^{2}\varepsilon_{\mathrm{\mathrm{N}}}^{2}m_{e}\varepsilon_{0}}{n_{pe}e^{2}\gamma}\right)^{1/4}  \label{EQ:MatchedR}  % EA 
\end{equation}

We %furthermore PM - EA ok
assume an initial normalized emittance of $\epsilon_{N} = 2\unit{\mu m}$. This emittance is possible to produce with a standard RF-injector, while at the same time yielding a sufficient narrow beam. % EA: comment, perhaps put a ref here [***ref].

Beam loading of a short witness beam is sensitive to its position relative to the electric field \cite{tzoufras:2009} %PM
as well as to its dephasing with respect to the wakefields when at low energy.
%PM A low gamma witness beam can easily drift out of phase with the drive beam wakefield. The drift of the beams is proportional to $\gamma^{-2}$,  
To eliminate %initial 
de-phasing of the witness beam, % PM
the initial beam energy %PMwas
is set such
that $\gamma_{eb} = \gamma_{pb} = 426.3$, giving an energy of $217\unit{MeV}$.  A lower initial energy is likely to be
sufficient for AWAKE Run 2 injection. 
% EA: we have studied this to some extent, but unfortunately we do not have a good refence for the last statement, no details in \cite{berglyd_olsen:2015}.

Eq. (\ref{EQ:MatchedR}) yields a transverse size of $\sigma_{x,y}$ of $5.25\unit{\mu m}$, narrow compared to the drive beam $\sigma_{x,y,proton} = 200\unit{\mu m}$.   The bunch length was set to $\sigma_{z}=60\unit{\mu m}$, based on earlier %PM
beam loading studies \cite{berglyd_olsen:2016}.  The charge %PM was
is adjusted to $100\unit{pC}$ for optimal beam loading, as discussed in the next section. We refer to the defined drive beam and witness beam parameter set as the \emph{base case}.  
% EA: the key figure, figure 2, was not discussed in the text until very far down!  I therefore added (text could possibly be improved)
Figure \ref{Fig:PlasmaDenTWake} shows the two beams from QuickPIC simulations of the initial time step, for the base case parameters; the proton bunch in blue, the trailing electron beam in red and the plasma electron density in grey.

%   EA: the discussing of charge could better be discussed in the "beam loading chapter" - the information has been moved to there
% the charge required to load the field is 
% density of the electron beam reaches that of the plasma density for a bunch charge of a few $\unit{pC}$.   
% but reaches optimal beam loading at around $100-200\unit{pC}$. 
%The implication here being that the witness beam produces its own non-linear
%wakefield already at the head of the beam. The majority of the electrons within it will therefore see a linear focusing
%force preventing the bulk of the beam from undergoing emittance growth.

% ******************************************************************************************************************** %
\subsection{Simulation parameters}\label{SIM}
% ******************************************************************************************************************** %

The relatively small size of the witness beam %PM compared to the proton beam put some restrictions on the transverse resolution of the witness beam as we need a simulation box wide enough to contain the larger beam while resolving the narrower one (see figure \ref{Fig:PlasmaDenTWake}). 
puts constraints on the transverse cell size and number in the simulations: small size to resolve the small electron beam and large number of to resolve the large proton beam and the wakefields. %
We use %PMd  - EA ok
a transverse grid cell size of $1.17\unit{\mu m}$, and of
$2.34\unit{\mu m}$ for the longitudinal grid cells for the simulations presented in section \ref{S:BL}. The witness beam
was simulated with $16.8\nexp{6}$ non-weighted particles.  ***NN*** plasma particles were simulated per grid cell.
%   EA: Veronica to fill in NN above

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamLoading}
    \caption{\label{Fig:BeamLoading} Top plot%PM plot shows the - EA ok
: unloaded longitudinal electric field with no witness beam
        (dashed blue line) and %PMthe 
        loaded field (whole blue line) along the beam axis. The beam density along the axis
        for both beams are shown in red. %PM The bottom plot compares the 
        Bottom plot: plasma densities along the beam axis for a drive
        beam with no witness beam (dashed green line), witness beam with no drive beam (dash-dotted green line), and
        both beams present (whole green line). $\xi = z - tc$ is the position in the simulation box, moving towards the
        left.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamPhaseSpace}
    \caption{\label{Fig:BeamPS} %PMThe p
    Phase space charge distribution of a $100\unit{pC}$, $60\unit{\mu m}$ long witness
        beam after $4\unit{m}$ of plasma. The mean momentum is $1.67\unit{GeV/c}$ with an RMS energy spread of
        $87\unit{MeV/c}$ ($5.2\%$) for the full beam.}
\end{figure}

@@

\begin{figure*}[hbt]
    \includegraphics[width=\linewidth,trim={0mm 0mm 0mm 0mm},clip]{figures/beamFilamentationAll}
    \caption{\label{Fig:BeamFilament} Plots \textbf{a} to \textbf{f} show the transverse phase space of the offset
        electron beam at different plasma positions. Plot \textbf{g} shows the macro particle mean position, and plot
        \textbf{h} their RMS spread. Plots \textbf{a}, \textbf{b} and \textbf{c}, as well as the blue lines in plots
        \textbf{g} and \textbf{h} represent particles %PM
        not in the ion column (see Fig.~\ref{Fig:BeamEmitt}), with position $1.40\unit{\mu m} < \xi < 1.42\unit{\mu m}$. Plots
        \textbf{d}, \textbf{e} and \textbf{f}, and the red lines in plots \textbf{g} and \textbf{h} represent particles %PM
        in the ion column 
        with position $1.55\unit{\mu m} < \xi < 1.57\unit{\mu m}$.}
\end{figure*}

@@

% EA: I think owned also by IST.  I will check with Weiming and Warren
The simulations for this study have been performed using the open source version of QuickPIC released in early 2017 and owned
by UCLA.
